---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# gghdr
 
Package `gghdr` imports the package [`hdrcde`](https://pkg.robjhyndman.com/hdrcde/) and provides tools for plotting highest density regions in the ggplot2 framework.

<!-- badges: start -->
<!-- badges: end -->

# Traditional boxplot

- central box bounded by Q1 and Q3 representing the interquartile range
- whiskers extending from Q1 - 1.5(Q3 - Q1)  to Q3 + to 1.5(Q3 - Q1), representing 99% coverage for large samples
- median represented by a horizontal line

```{r boxplot, echo=FALSE}
library(hdrcde)
library(ggplot2)
ggplot(faithful, aes(y=eruptions)) +  geom_boxplot()
```

# hdr.boxplot (Existing)

There are different ways to summarize a distribution but the highest density region allows the display of multimodality.

- Region bounded by the interquartile range is replaced by 50% HDR
- Region bounded by the whiskers is replaced by the 99% HDR
- The mode is represented by a horizontal line
 
 <!-- In both HDR and box plots, the interquartile range or 50% HDR will have a coverage probability of 50%. -->


```{r hdrcde-boxplot, echo=TRUE}
library(hdrcde)
hdr.boxplot(faithful$eruptions)
```

<!-- # ```{r hdrcde-boxplot_more, echo=TRUE, eval = FALSE} -->
<!-- # hdr.boxplot(x, prob = c(99, 50), h = hdrbw(BoxCox(x, lambda), -->
<!-- #   mean(prob)), lambda = 1, boxlabels = "", col = gray((9:1)/10), -->
<!-- #   main = "", xlab = "", ylab = "", pch = 1, border = 1, -->
<!-- #   outline = TRUE, space = 0.25, ...) -->
<!-- # ``` -->

# geom_boxplot_hdr (Proposal)

```{r gg_hdr-boxplot, echo=TRUE, eval = FALSE}
library(ggplot2)
faithful %>% ggplot(aes(y = )) +  geom_hdr_boxplot()
```


# hdr.den (Existing)

- a density plot
- the endpoints of each interval in each HDR on the x-axis
- a straight line showing value of the density at the boundaries of each HDR

ˆ
```{r hdr.den}
hdr.den(faithful$eruptions)
#ggplot(faithful, aes(x=eruptions)) +  geom_density()
```

# gg_hdr (Proposal)

- geom_density() - to represent the density
- geom_hdr_rug() - The endpoints of each interval in each HDR on the x-axis
- geom_hline() - The value of the density at the boundaries of each HDR

gghdr() =  geom_density() + geom_hdr_rug() + geom_hline()


```{r gg_hdr, echo=TRUE, eval = FALSE}
library(ggplot2)
faithful %>% ggplot(aes(y = , x = )) +  geom_hdr()
```


